Misexpression of germline programs with loss of the X-linked intellectual disability factor KDM5C

### 2024-04-18 submission/manuscript.Rmd
bullet point draft of EpiLC data

>>>>

## _Kdm5c_-KO cells express master regulators of germline identity in an _in vitro_ model of early embryogenesis

Misexpression of germline genes in the _Kdm5c_-KO brain suggests mutants fail to demarcate germline and somatic cellular identity. Germ cells are typically distinguished from somatic cells soon after the embryo implants into the uterine wall[@magnusdottirHowMakePrimordial2014; @gunesdoganPrimordialGermCell2014] when a subset of epiblast stem cells become the primordial germ cells (PGCs) while the remainder differentiate into the ectoderm, mesoderm, and endoderm to form the body's somatic tissues[@bardotMouseGastrulationCoordination2020]. This developmental time point can be modeled _in vitro_ through differentiation of embryonic stem cells (ESCs) into post-implantation epiblast-like stem cells (EpiLCs) (Figure 3A, top). Previous studies have demonstrated that while some germline-enriched genes are also expressed in embryonic stem cells (ESCs), they are silenced as they differentiate into EpiLCs[@mochizukiRepressionGermlineGenes2021]. Therefore, to elucidate the mechanism underlying aberrant germline genes expression in the mature _Kdm5c_-KO brain, we first assessed their expression in male _Kdm5c_-KO EpiLCs using our previously generated RNA-seq dataset[@samantaActivationXistEvolutionarily2022].

* We did not observe any apparent morphology changes between wild-type and _Kdm5c_-KO cells during ESC to EpiLC differentiation (Figure 3A, bottom). Additionally, _Kdm5c_-KO EpiLCs downregulated ESC pluripotency genes and upregulated EpiLC differentiation genes similar to wild-type (Figure 3B).

* We then identified _Kdm5c_-KO EpiLC DEGs through DESeq2[@loveModeratedEstimationFold2014a] (log2 fold change > `r l2fcco`, q < 0.1) and observed widespread dysregulation of tissue-enriched genes, with the largest number of genes belonging to the brain and testis (Figure 3C).

* Of these testis genes, we also observed misexpression of germline-enriched genes, including _Cyct_ (Figure 3D)

* To evaluate whether germline mRNAs are constitutively dysregulated or change over the course of development, we compared EpiLC germline DEGs to those expressed in the _Kdm5c_-KO brain. The majority of EpiLC germline DEGs were unique to EpiLCs, with only a few shared across brain sequencing datasets (Figure 3E). Only two germline genes were consistently misexpressed in all datasets, _DNA segment, Chr 1, Pasteur Institute 1 (D1Pas1)_ and _Testis expressed gene 14 (Tex14)_.

* We then compared the typical function of brain and EpiLC germline DEGs and found EpiLC germline DEGs displayed strong enrichment of meiosis-related gene ontologies, including XXXXX (Figure 3F).

* Surprisingly, germline DEGs unique to _Kdm5c_-KO EpiLCs included master regulators of germline identity, such as _Stimulated by retinoic acid 8 (Stra8)_ and _Deleted in azoospermia like (Dazl)_ (Figure 3G). These genes are typically expressed during embryonic germ cell development to commit PGCs to the germline fate, but they are also expressed later in life to trigger meiotic gene expression programs[@koubovaRetinoicAcidRegulates2006; @linGermCellIntrinsicExtrinsic2008; @endoRetinoicAcidGerm2019].

* _Dazl_ is a translational regulator essential for germ cell function[@liDAZLMasterTranslational2019]. Interestingly, a significant portion of germline transcripts expressed in _Kdm5c_-KO EpiLCs are known binding targets of DAZL, including _Stra8_[@mikedisDAZLMediatesBroad2020] (XY _Kdm5c_-KO EpiLCs p = 1.698e-07, Fisher’s Exact Test). This suggests expression of DAZL protein would enable the translation of other aberrant germline transcripts, influencing their ability to impact on _Kdm5c_ mutant cellular function.

* Therefore, we then stained _Kdm5c_-KO EpiLCs for DAZL protein expression through immunocytochemistry (Figure 3H). We found about 25% of _Kdm5c_-KO EpiLCs expressed DAZL protein in their cytoplasm (p = 0.0015, Welch's t-test, XY KO versus WT), consistent with the subcellular localization observed when DAZL is stained in spermatogonia[@mikedisDAZLMediatesBroad2020].
